\textbf{Effective application of machine learning to a microbiome-based classification problem}

### 2020-01-20 submission/manuscript.tex
update for rebuttal

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Effective application of machine learning to a microbiome-based classification problem},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

\title{\textbf{Effective application of machine learning to a microbiome-based
classification problem}}
\author{}
\date{\vspace{-2.5em}}

@@

\textbf{Interpretation of each ML model.} Interpretability is related to
the degree to which humans can understand the reasons behind a model
prediction (34--36). Because we often use ML models not just to predict
a health outcome, but also to identify potential biomarkers for disease,
model interpretation becomes crucial for microbiome studies. ML models
often decrease in interpretability as they increase in complexity. In
this study, we used two methods to help interpret our models.

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU was associated with classifying a subject as
being healthy or having an SRN. It was encouraging that many of the
highest-ranked OTUs were shared across these three models (e.g., OTUs
50, 426, 609, 822, 1239). The benefit of this approach was that the
results of the analysis were based on the trained model parameters and
provided information regarding the sign and magnitude of the impact of
each OTU. However, this approach is limited to linear models or models
with prespecified interaction terms.

@@

@@

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures that create collinearity in the datasets. This can hinder our
ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (38). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we partially accounted for collinearity by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other; however, we could reduce
the correlation threshold to further investigate the relationships among
correlated features. It is important to know the correlation structures
of the data to avoid misinterpreting the models. This is likely to be a
particular problem with shotgun metagenomic datasets where collinearity
will be more pronounced due to many genes being correlated with one
another because they come from the same chromosome. To identify the true
underlying microbial factors of a disease, it is crucial to follow up on
any correlation analyses with further hypothesis testing and
experimentation for biological validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with healthy
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with healthy colons and those with SRNs.
Furthermore, as the number of classes increases, more samples are
required for each category to train an accurate model. We also did not
use regression-based analyses to predict a non-categorical outcome. We
have previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (39). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (12, 40, 41), but microbiome datasets
often suffer from having many features and small sample sizes, which
result in overfitting.

Our framework provides a reproducible structure to investigators wanting
to train, evaluate, and interpret their own ML models to generate
hypotheses regarding which OTUs might be biologically relevant. However,
deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly more challenging and distinct undertaking
(42). For example, we currently lack standardized methods to collect
patient samples, generate sequence data, and report clinical data. We
are also challenged by the practical constraints of OTU-based
approaches. The de novo algorithms commonly in use are slow, require
considerable memory, and result in different OTU assignments as new data
are added (43). Finally, we also need independent validation cohorts to
test the performance of a diagnostic model. To realize the potential for
using ML approaches with microbiome data, it is necessary that we direct
our efforts to overcome these challenges.

@@

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with healthy
colons and those with screen relevant neoplasias (SRNs). The healthy
class included patients with healthy colons or non-advanced adenomas,
whereas the SRN class included patients with advanced adenomas or
carcinomas (44). Patients with an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas (44). There were 172 patients
with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
Of the 198 adenomas, 109 were identified as advanced adenomas. Together
261 patients were classified as healthy and 229 patients were classified
as having an SRN.

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
generated by Sze et al (45), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (43, 46, 47);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1, and the
lowest would be 0.

@@

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter.For
L2-regularized logistic regression, L1 and L2-regularized SVM with
linear and radial basis function kernels, we tuned the cost
hyperparameter, which controls the regularization strength, where
smaller values specify stronger regularization. For SVM with radial
basis function kernel, we also tuned the sigma hyperparameter, which
determines the reach of a single training instance where, for a high
value of sigma, the SVM decision boundary will be dependent on the
points that are closest to the decision boundary. For the decision tree
model, we tuned the depth of the tree where the deeper the tree, the
more splits it has. For random forest, we tuned the number of features
to consider when looking for the best tree split. For XGBoost, we tuned
the learning rate and the fraction of samples used for fitting the
individual base learners. Performing a grid search for hyperparameter
selection might not be feasible for when there are more than two
hyperparameters to tune for. In such cases, it is more efficient to use
random search or recently developed tools such as Hyperband to identify
good hyperparameter configurations (48).

@@

\hypertarget{ref-sze_fecal_2019}{}
39. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
\textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
acids are not predictive of colonic tumor status and cannot be predicted
based on bacterial community structure. mBio \textbf{10}:e01454--19.
doi:\href{https://doi.org/10.1128/mBio.01454-19}{10.1128/mBio.01454-19}.

\hypertarget{ref-kocheturov_massive_2019}{}
40. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
2019. Massive datasets and machine learning for computational
biomedicine: Trends and challenges. Ann Oper Res \textbf{276}:5--34.
doi:\href{https://doi.org/10.1007/s10479-018-2891-2}{10.1007/s10479-018-2891-2}.

\hypertarget{ref-kim_improved_2018}{}
41. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
method for prediction of cancer prognosis by network learning. Genes
\textbf{9}:478.
doi:\href{https://doi.org/10.3390/genes9100478}{10.3390/genes9100478}.

\hypertarget{ref-wiens_no_2019}{}
42. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
\textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
\textbf{Jung K}, \textbf{Heller K}, \textbf{Kale D}, \textbf{Saeed M},
\textbf{Ossorio PN}, \textbf{Thadaney-Israni S}, \textbf{Goldenberg A}.
2019. Do no harm: A roadmap for responsible machine learning for health
care. Nat Med \textbf{25}:1337--1340.
doi:\href{https://doi.org/10.1038/s41591-019-0548-6}{10.1038/s41591-019-0548-6}.

\hypertarget{ref-westcott_opticlust_2017}{}
43. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
Improved Method for Assigning Amplicon-Based Sequence Data to
Operational Taxonomic Units. mSphere \textbf{2}.
doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.

\hypertarget{ref-redwood_stool_2016}{}
44. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
\textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
\textbf{Tiesinga JJ}, \textbf{Devens ME}, \textbf{Alberts SR},
\textbf{Mahoney DW}, \textbf{Yab TC}, \textbf{Foote PH}, \textbf{Smyrk
TC}, \textbf{Provost EM}, \textbf{Ahlquist DA}. 2016. Stool DNA testing
for screening detection of colorectal neoplasia in alaska native people.
Mayo Clin Proc \textbf{91}:61--70.
doi:\href{https://doi.org/10.1016/j.mayocp.2015.10.008}{10.1016/j.mayocp.2015.10.008}.

\hypertarget{ref-sze_leveraging_2018}{}
45. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\hypertarget{ref-schloss_introducing_2009}{}
46. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.

\hypertarget{ref-rognes_vsearch_2016}{}
47. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.

\hypertarget{ref-li_hyperband:_2016}{}
48. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
\textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
bandit-based approach to hyperparameter optimization. arXiv:160306560
{[}cs, stat{]}.

@@

\textbf{Figure S5. Interpretation of the linear ML models with
permutation importance.} (A) L1-regularized SVM with linear kernel, (B)
L2-regularized SVM with linear kernel, and (C) L2-regularized logistic
regression were interpreted using permutation importance using held-out
test set.

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Effective application of machine learning to a microbiome-based classification problem},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{\textbf{Effective application of machine learning to a microbiome-based
classification problem}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

@@

\textbf{Interpretation of each ML model.} To evaluate the overall
structure of the CRC dataset and show that the structure of the human
microbiome is remarkably variable among individuals with SRNs, we
analyzed the Bray-Curtis distances with an NMDS ordination {[}Figure
S5{]}. The lack of seperation between healthy patients and patients with
SRNs, show that traditional statistical approaches do not help us to
adequately address disease ecology. This is why we often use ML models
not just to predict a health outcome, but also to identify potential
biomarkers for disease. Therefore, model interpretation becomes crucial
for microbiome studies. Interpretability is related to the degree to
which humans can understand the reasons behind a model prediction
(34--36). ML models often decrease in interpretability as they increase
in complexity. In this study, we used two methods to help interpret our
models.

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU was associated with classifying a subject as
being healthy or having an SRN. It was encouraging that many of the
highest-ranked OTUs were shared across these three models (e.g., OTUs
50, 426, 609, 822, 1239). The great benefit of this approach was knowing
the sign and magnitude of each OTU coefficient in the trained model.
This allowed us to immidiately interpret negative and positive
coefficient signs as protective and risk factors, respectively and the
magnitude as the impact of these factors. However, this approach is
limited to linear models or models with prespecified interaction terms.

@@

We also compared the top 20 OTUs selected by permutation importance in
L2-regularized logistic regression and the highest performing tree-based
models, random forest and XGBoost. 2 and 5 OTUs, respectively, were
shared among the models. These results indicate that we were able to
identify a few more OTUs that are deemed important by both linear and
non-linear models when we used permutation importance as our
interpretation method.

@@

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures that create collinearity in the datasets. This can hinder our
ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (38). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we partially accounted for collinearity by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other; however, we could reduce
the correlation threshold to further investigate the relationships among
correlated features. We are choosing to leave the decision to define
correlation and setting up correlations thresholds to the researchers
for their own microbiome questions. However, we would like to urge them
to investigate the correlation structures of the data to avoid
misinterpreting the models. This is likely to be a particular problem
with shotgun metagenomic datasets where collinearity will be more
pronounced due to many genes being correlated with one another because
they come from the same chromosome. True causal mechanisms (e.g.~role of
microbiome in colorectal cancer) cannot be explained solely by the
highest performing machine learning model (39). To identify the true
underlying microbial factors of a disease, it is crucial to follow up on
any correlation analyses with further hypothesis testing and
experimentation for biological validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with healthy
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with healthy colons and those with SRNs.
Furthermore, as the number of classes increases, more samples are
required for each category to train an accurate model. We also did not
use regression-based analyses to predict a non-categorical outcome. We
have previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (40). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (12, 41, 42), but microbiome datasets
often suffer from having many features and small sample sizes, which
result in overfitting.

Our framework provides a reproducible structure to investigators wanting
to train, evaluate, and interpret their own ML models to generate
hypotheses regarding which OTUs might be biologically relevant. However,
deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly more challenging and distinct undertaking
(43). For example, we currently lack standardized methods to collect
patient samples, generate sequence data, and report clinical data. We
are also challenged by the practical constraints of OTU-based
approaches. The de novo algorithms commonly in use are slow, require
considerable memory, and result in different OTU assignments as new data
are added (44). Finally, we also need independent validation cohorts to
test the performance of a diagnostic model. To realize the potential for
using ML approaches with microbiome data, it is necessary that we direct
our efforts to overcome these challenges.

@@

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with healthy
colons and those with screen relevant neoplasias (SRNs). The healthy
class included patients with healthy colons or non-advanced adenomas,
whereas the SRN class included patients with advanced adenomas or
carcinomas (45). Patients with an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas (45). There were 172 patients
with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
Of the 198 adenomas, 109 were identified as advanced adenomas. Together
261 patients were classified as healthy and 229 patients were classified
as having an SRN.

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
generated by Sze et al (46), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (44, 47, 48);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1, and the
lowest would be 0.

@@

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter.For
L2-regularized logistic regression, L1 and L2-regularized SVM with
linear and radial basis function kernels, we tuned the cost
hyperparameter, which controls the regularization strength, where
smaller values specify stronger regularization. For SVM with radial
basis function kernel, we also tuned the sigma hyperparameter, which
determines the reach of a single training instance where, for a high
value of sigma, the SVM decision boundary will be dependent on the
points that are closest to the decision boundary. For the decision tree
model, we tuned the depth of the tree where the deeper the tree, the
more splits it has. For random forest, we tuned the number of features
to consider when looking for the best tree split. For XGBoost, we tuned
the learning rate and the fraction of samples used for fitting the
individual base learners. Performing a grid search for hyperparameter
selection might not be feasible for when there are more than two
hyperparameters to tune for. In such cases, it is more efficient to use
random search or recently developed tools such as Hyperband to identify
good hyperparameter configurations (49).

@@

\hypertarget{ref-li_accurate_2020}{}
39. \textbf{Li J}, \textbf{Liu L}, \textbf{Le TD}, \textbf{Liu J}. 2020.
Accurate data-driven prediction does not mean high reproducibility. Nat
Mach Intell \textbf{2}:13--15.
doi:\href{https://doi.org/10.1038/s42256-019-0140-2}{10.1038/s42256-019-0140-2}.

\hypertarget{ref-sze_fecal_2019}{}
40. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
\textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
acids are not predictive of colonic tumor status and cannot be predicted
based on bacterial community structure. mBio \textbf{10}:e01454--19.
doi:\href{https://doi.org/10.1128/mBio.01454-19}{10.1128/mBio.01454-19}.

\hypertarget{ref-kocheturov_massive_2019}{}
41. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
2019. Massive datasets and machine learning for computational
biomedicine: Trends and challenges. Ann Oper Res \textbf{276}:5--34.
doi:\href{https://doi.org/10.1007/s10479-018-2891-2}{10.1007/s10479-018-2891-2}.

\hypertarget{ref-kim_improved_2018}{}
42. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
method for prediction of cancer prognosis by network learning. Genes
\textbf{9}:478.
doi:\href{https://doi.org/10.3390/genes9100478}{10.3390/genes9100478}.

\hypertarget{ref-wiens_no_2019}{}
43. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
\textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
\textbf{Jung K}, \textbf{Heller K}, \textbf{Kale D}, \textbf{Saeed M},
\textbf{Ossorio PN}, \textbf{Thadaney-Israni S}, \textbf{Goldenberg A}.
2019. Do no harm: A roadmap for responsible machine learning for health
care. Nat Med \textbf{25}:1337--1340.
doi:\href{https://doi.org/10.1038/s41591-019-0548-6}{10.1038/s41591-019-0548-6}.

\hypertarget{ref-westcott_opticlust_2017}{}
44. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
Improved Method for Assigning Amplicon-Based Sequence Data to
Operational Taxonomic Units. mSphere \textbf{2}.
doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.

\hypertarget{ref-redwood_stool_2016}{}
45. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
\textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
\textbf{Tiesinga JJ}, \textbf{Devens ME}, \textbf{Alberts SR},
\textbf{Mahoney DW}, \textbf{Yab TC}, \textbf{Foote PH}, \textbf{Smyrk
TC}, \textbf{Provost EM}, \textbf{Ahlquist DA}. 2016. Stool DNA testing
for screening detection of colorectal neoplasia in alaska native people.
Mayo Clin Proc \textbf{91}:61--70.
doi:\href{https://doi.org/10.1016/j.mayocp.2015.10.008}{10.1016/j.mayocp.2015.10.008}.

\hypertarget{ref-sze_leveraging_2018}{}
46. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\hypertarget{ref-schloss_introducing_2009}{}
47. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.

\hypertarget{ref-rognes_vsearch_2016}{}
48. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.

\hypertarget{ref-li_hyperband:_2016}{}
49. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
\textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
bandit-based approach to hyperparameter optimization. arXiv:160306560
{[}cs, stat{]}.

@@

\textbf{Figure S5. NMDS ordination with Bray-Curtis distances.}

\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}

\textbf{Figure S6. Interpretation of the linear ML models with
permutation importance.} (A) L1-regularized SVM with linear kernel, (B)
L2-regularized SVM with linear kernel, and (C) L2-regularized logistic
regression were interpreted using permutation importance using held-out
test set.
